Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-09 survey/survey.tex
_

Instead of spaces consisting solely of unstructured isolated points,
we now investigate structured spaces built from sums and products of
elementary spaces. This structure corresponds to the building blocks
of type theory which are: the empty type ($\bot$), the unit type
($\top$), the sum type ($\uplus$), and the product ($\times$)
type. Before getting into the formal theory, let's consider possible
deformations on the space
$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
product of two subspaces: the subspace $(\top \uplus \bot)$ which
itself is the sum of the space~$\top$ containing one element
$\texttt{tt}$ and the empty space $\bot$ and the subspace
$(\top \uplus \top)$ which is the sum of two spaces each containing
the one element $\texttt{tt}$. First, as discussed in the previous
section, any deformation of this space must at least preserve the
number of elements: we can neither create nor destroy points during
any continuous deformation. Seeing that the number of elements in our
example space is 2, a reasonable hypothesis is that we can deform the
space above to any other space with 2 elements such as
$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line, change the orientation of the grid by exchanging
the axes, as well as do other transformations to be discussed in this
section that preserve the number of points.

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

\begin{defn}[Homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
\end{defn}

@@

We are now in position to formally state the appropriate commutative
semiring structure for types. After replacing the set $R$ with a
universe $U$, we also replace the algebraic use of $=$ in
Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
this change, we can indeed prove that types (with
$\bot, \top, \uplus, \times$) form a commutative semiring. The reader
familiar with universal algebra should pause and ponder a bit about
what we have done. We have lifted \emph{equality} from being in the
signature of the ambient logic and instead put it in the signature of
the algebraic structure of interest. Then, and only then, are we in a
solid position to show that we have a concept (commutative semiring)
which is flexible enough for types to be an instance.

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first has \emph{two} proofs of inhabitation
($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's daemon: even though the data is the same, they are tagged
differently, and this data is indeed information, and must be
preserved.

>>>>

\roshan{Let's not substitute space for set without first motivating
  the use of the word.}  Instead of spaces consisting solely of
unstructured isolated points, we now investigate structured spaces
built from sums and products of elementary spaces. This structure
corresponds to the building blocks of type theory which are: the empty
type ($\bot$), the unit type ($\top$), the sum type ($\uplus$), and
the product ($\times$) type. Before getting into the formal theory,
let's consider possible deformations on the space $(\top \uplus \bot)
\times (\top \uplus \top)$. This space is the product of two
subspaces: the subspace $(\top \uplus \bot)$ which itself is the sum
of the space~$\top$ containing one element $\texttt{tt}$ and the empty
space $\bot$ and the subspace $(\top \uplus \top)$ which is the sum of
two spaces each containing the one element $\texttt{tt}$. First, as
discussed in the previous section, any deformation of this space must
at least preserve the number of elements: we can neither create nor
destroy points during any continuous deformation. Seeing that the
number of elements in our example space is 2, a reasonable hypothesis
is that we can deform the space above to any other space with 2
elements such as $\top \uplus \top$ or $\top \uplus (\top \uplus
\bot)$. What this really means is that we are treating the sum and
product structure as malleable. For example, imagining a product
structure as arranged in a grid, by stretching we can turn this
structure to a sum structure arranged in a line, change the
orientation of the grid by exchanging the axes, as well as do other
transformations to be discussed in this section that preserve the
number of points.

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set.''  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
type $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal} such that a term exhibiting this type can
then be shown in the formal system. This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except in univalent type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into $B \times
A$. Which motivates the introduction of type \emph{equivalences}. To
do this, we need a few important auxiliary concepts.

\roshan{Let's define $\equiv$ below or provide a readable
  reference. We should not assume that its meaning is automatically
  understood.}
\begin{defn}[Homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
\end{defn}

@@

We are now in position to formally state the appropriate commutative
semiring structure for types. After replacing the set $R$ with a
universe $U$, we also replace the algebraic use of $=$ in
Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
this change, we can indeed prove that types (with $\bot, \top, \uplus,
\times$) form a commutative semiring. The reader familiar with
universal algebra should pause and ponder a bit about what we have
done. We have lifted \emph{equality} from being in the signature of
the ambient logic and instead put it in the signature of the algebraic
structure of interest. \roshan{The last coupe of sentences are
  unclear. Let's strive to keep things easy to understand. Can we
  spell them out the tradeoffs here and give an example or two to
  clarify the point. If this is indeed an important aspect of things
  let's spend some space explaining it.} Then, and only then, are we
in a solid position to show that we have a concept (commutative
semiring) which is flexible enough for types to be an instance.

@@

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem.'' In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first can have \emph{two} proofs of inhabitation
($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's daemon: even though the data is the same, they are tagged
differently, and this data is indeed information, and must be
preserved.
